Density-dependent selection and the limits of relative fitness

### 2018-04-23 manuscript.tex
l

For haploids tracked in discrete time, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}. In the particular case that two types are present, the selection equation takes the familiar form
\begin{equation}
\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$. Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would then be relative log fitnesses).

In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use the continuous time Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic growth rates. However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density. Thus, assuming viability, the increase in population density must be checked by crowding, implying that the Malthusian parameters $r_i$ eventually decline to zero (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

>>>>

For haploids tracked in discrete time, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''. The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$. In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}. In the particular case that two types are present, the discrete and continuous changes in frequency can be written respectively as $\Delta p_i=\frac{W_i-W_j}{\overline{W}} p_i p_j$ and $\frac{d p_i}{dt}=s p_i p_j$
\begin{equation}
\frac{d p_i}{dt}=s p_i p_j \label{eq:canonical}
\end{equation}
where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.

Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$. These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp. 468]{barton_2007}. Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would then be relative log fitnesses).

Henceforth our point of comparison we will be the continuous time Eq.~\eqref{eq:canonical}.

In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate. The selection coefficient $s$ is then simply the difference in intrinsic growth rates. However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density. Thus, assuming viability, the increase in population density must be checked by crowding, implying that the Malthusian parameters $r_i$ eventually decline to zero (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.
